# MetaData {#00}

### Source code {-}

Bookdown source code: [PortofolioBookdown](https://github.com/ThijmenWeijgertze/PortofolioBookdown)

Website source code: [ThijmenWeijgertze-portofolio](https://github.com/ThijmenWeijgertze/ThijmenWeijgertze-portfolio)

Website link: [thijmenweijgertze.github.io/ThijmenWeijgertze-portfolio/](https://thijmenweijgertze.github.io/ThijmenWeijgertze-portfolio/)

### Dependency packages {-}

-   devtools
-   tidyverse
-   tidymodels
-   bookdown
-   reactable
-   gt
-   fs
-   tidyverse
-   RColorBrewer
-   ranger
-   readxl
-   ggfortify
-   tableone
-   broom.mixed
-   dotwhisker (dependencies: leeper/prediction & leeper/margins)
-   twPackage (devtools::install_github("ThijmenWeijgertze/twPackage", build_vignettes = TRUE))

Installing dependency packages (if it doesn't download automatically please install the packages by hand):
```{r dependency packages}
# if not installed: installing dependency packages from cran
for (package in c("devtools", "tidyverse", "tidymodels", "bookdown", "reactable", "gt", "fs", "RColorBrewer", "readxl", "ggfortify")) {
  if (!require(package, character.only = TRUE)) {
    install.packages(package, dependencies = TRUE)
    library(package, character.only = TRUE)
  } else {
    library(package, character.only = TRUE)
  }
}

# install twPackage from github
if (!require("twPackage", character.only = TRUE)) {
  devtools::install_github("ThijmenWeijgertze/twPackage", build_vignettes = TRUE)
  library(package, character.only = TRUE)
}

# prediction dependency for dotwhisker
if (!require("prediction", character.only = TRUE)) {
    remotes::install_github("leeper/prediction")
    library("prediction", character.only = TRUE)
  } else {
    library("prediction", character.only = TRUE)
  }

# margins dependency for dotwhisker
if (!require("margins", character.only = TRUE)) {
    remotes::install_github("leeper/margins")
    library("margins", character.only = TRUE)
  } else {
    library("margins", character.only = TRUE)
  }

# Dotwhisker itself
if (!require("dotwhisker", character.only = TRUE)) {
    remotes::install_github("fsolt/dotwhisker")
    library("dotwhisker", character.only = TRUE)
  } else {
    library("dotwhisker", character.only = TRUE)
  }
```

### Directory roadmap {-}

The structure of this portfolio's source code is as follows:

```{r directory roadmap, warning=FALSE, message=FALSE}
library(fs)
library(tidyverse)

dir_tree(".", recurse = TRUE, regexp = "^.gitignore$|^[^_|(.git)]", all = TRUE)
```

### dataRaw & data folder {-}

<p class="justify">The dataRaw folder holds the original data files. These data files are unedited and remain as untouched in the dataRaw folder as back-up. The data within the data folder are used within the Rmarkdown files and may differ from the original source in dataRaw.<p> 

The data folder consists of the following files:
- CE-LIQ-FLOW-062_Tidydata.xlsx: C.elegans data used in the C.elegans plate analysis
- GuerrillaAnalytics.png: a screenshot of a project's directory tree to demonstrate the guerrilla analytics principles 
