# MetaData {#00}

### Source code {-}

Bookdown source code: [PortofolioBookdown](https://github.com/ThijmenWeijgertze/PortofolioBookdown)

Website source code: [ThijmenWeijgertze-portofolio](https://github.com/ThijmenWeijgertze/ThijmenWeijgertze-portfolio)

twPackage source code: [twPackage](https://github.com/ThijmenWeijgertze/twPackage)

Website link: [thijmenweijgertze.github.io/ThijmenWeijgertze-portfolio/](https://thijmenweijgertze.github.io/ThijmenWeijgertze-portfolio/)

### How to build the book localy {-}

- Download the rmarkdown package (cran)
- Download the bookdown package (cran)
- Download the bookdown source code
- Open the Rproject file
- Run the "Installing dependency packages" code chunk
- Press the build book button
- If there is no build book button use: `bookdown::render_book(".")`
- If something goes wrong try to install the dependencies manually

### Dependency packages {-}

-   devtools
-   tidyverse
-   tidymodels
-   bookdown
-   reactable
-   gt
-   fs
-   tidyverse
-   RColorBrewer
-   ranger
-   readxl
-   ggfortify
-   tableone
-   randomForest
-   broom.mixed
-   rpart.plot
-   vip
-   MASS
-   twPackage (devtools::install_github("ThijmenWeijgertze/twPackage", build_vignettes = TRUE))

Installing dependency packages:
```{r Installing dependency packages}
# if not installed: installing dependency packages from cran
for (package in c("devtools", "tidyverse", "tidymodels", "bookdown", "reactable", "gt", "fs", "tidyverse", "RColorBrewer", "ranger", "readxl", "ggfortify", "tableone", "randomForest", "broom.mixed", "rpart.plot", "vip", "MASS")) {
  if (!require(package, character.only = TRUE)) {
    install.packages(package, dependencies = TRUE)
    library(package, character.only = TRUE)
  } else {
    library(package, character.only = TRUE)
  }
}

# install twPackage from github
if (!require("twPackage", character.only = TRUE)) {
  devtools::install_github("ThijmenWeijgertze/twPackage", build_vignettes = TRUE)
  library(package, character.only = TRUE)
}
```

### Directory roadmap {-}

The structure of this portfolio's source code is as follows:

```{r directory roadmap, warning=FALSE, message=FALSE}
library(fs)
library(tidyverse)

dir_tree(".", recurse = TRUE, regexp = "^.gitignore$|^[^_|(.git)]", all = TRUE)
```

### Project directory folder

- .gitignore: file to withold files from being pushed to github
- 00-METADATA.Rmd: contains the metadata of the project
- 01_CV.Rmd: contains the resume
- 02-GuerillaAnalytics.Rmd: contains the folder structure of this project and an older project to show of guerilla analytics principles
- 03-CElegans.Rmd: contains the C.elegans plate analysis
- 04-ReproducibleResearch.Rmd: contains both the open peer review and the reproduced code
- 05-MachineLearning.Rmd: contains the breast cancer data analysis using random forest alghorithms from tidymodels
- 06-Package.Rmd: contains a selfmade r package called twPackage
- 07-DataScienceProject.Rmd: contains the information about the data science project AMR
- 08-CovidCases.Rmd: contains the covid cases analysis
- 19-appendices.Rmd: contains the appendix with the reproduced COVID-19 vaccin paper code
- 20-SessionInfo.Rmd: contains the session info
- 21-Bibliography.Rmd: contains the references
- css folder: contains the style.css file for the website (only works if it's in the css folder)
- data folder: contains the data used in the code
- dataRaw folder: contains the original untouched data files
- index.Rmd: contains the index page (the page from where the bookdown starts)
- LICENSE: contains the MIT license
- PortofolioBookdown.Rproj: contains the Rproject 
- README.md: redirects to the repositories and this METADATA file
- References.bib: contains the references used
- scripts folder: contains the scripts used in the code

### css folder {-}

- style.css: consists of the css code for the bookdown layout

### dataRaw & data folder {-}

<p class="justify">The dataRaw folder holds the original data files. These data files are unedited and remain as untouched in the dataRaw folder as back-up. The data within the data folder are used within the Rmarkdown files and may differ from the original source in dataRaw.<p> 

The data & dataRawfolder folder consists of the following files:

- CE-LIQ-FLOW-062_Tidydata.xlsx: C.elegans data used in the C.elegans plate analysis @louterCELIQFLOW
- GuerrillaAnalytics.png: a screenshot of a project's directory tree to demonstrate the guerrilla analytics principles 
- education.png, age.png, caseperm.png, supplementary.png: screenshots used for reproducing r code (chapter open peer review)
- resampling.png: illustration of what resampling does in machine learning context
- covidData.csv: Covid data used in the Covid cases analysis @DataDailyNumber2022
- osfstorage-archive: contains the rmd of the reproduced code @palayewCOVID19VaccinePaper2020
- breast-cancer.data: contains the breast cancer data for the random forest algorithm chapter @matjazzwitterBreastCancer1988
- breast-cancer.names: contains the breast cancer data's metadata @matjazzwitterBreastCancer1988
- dice.Rds and dice.csv: illustrate the tw_csv_rds function from the twPackage. Which saves a tibble as a csv and rds file
- ThijmenWeijgertzeResume.pdf: My resume in pdf format voor the resume chapter

### scripts {-}

- cleanBook.R: script to use the clean book function to clean up bookdown files after building the book. Usage: source("data/CleanBook.R")
- UpdateBib(DoesntWork).R: should have been used to update the References.bib file, but I could not get it working. I recommend ignoring this script
