# Rpackage (twPackage) {#06}

### Introduction

The package consists of 4 functions for a DataScience portofolio. The package is created with the following guide: @handleywickhamPackages2ndEdition2023. 

### Installation

The package can be installed via github: 
- devtools::install_github("ThijmenWeijgertze/twPackage", build_vignettes = TRUE)

The twPackage makes use of the following packages
- here
- reactable
- readr
- dplyr
- magrittr

Suggested packages: 
- knitr
- rmarkdown

### Function and data explanation

tw_factor()

-   description: This command changes a column type to factor with unique levels using the unique() command
-   usage: tw_factor(x)
-   x: The column to be changed into a factor
-   returns: Column with type factor

tw_csv_rds()

-   description: This command exports a variable containing a dataframe or tibble into a .csv and .rds file at a chosen location. The packages relies on the readr and here package. A project must be loaded in order to make this command work
-   usage: tw_csv_rds(x, path)
-   x Dataframe or tibble
-   path The path where the files must be stored. The path must be specified starting at the rproject directory. Do not add .csv or .rds at the end of the file name
-   returns: A .csv and .tsv from a dataframe or tibble at the chosen location

tw_table()

-   description: This command creates a basic table with the reactable package.
-   usage: tw_table(x)
-   x: A variable containing a dataframe or tibble
-   returns: A basic table made with the reactable package

tw_filter_select()

-   description: This command creates a basic table with the reactable package.
-   usage: tw_filter_select(x, filter_col, filter_value, select)
-   x: The table to be filtered
-   filter_col: The column where the filter_value can be found
-   filter_value: The value to be filtered on
-   select: The columns to be selected
-   returns: A filtered and selected data

Pokemon data

-   description: A dataframe with 800 rows and 13 columns containing pokemon data from around 2016.
-   usage: pokemon
-   returns: The pokemon dataset
-   source: [Pokemon data](https://gist.github.com/armgilles/194bcff35001e7eb53a2a8b441e8b2c6)

### Function examples

loading package

```{r load tw package}
# load package
library(twPackage)
```

tw_factor()

```{r tw_factor}
# Changing column Type.1 into a factor with unique levels
pokemon$Type.1 <- tw_factor(pokemon$Type.1)
levels(pokemon$Type.1)
```

tw_csv_rds()

```{r tw_csv_rds}
# Creating a table with dice data
dice_table <- tibble(
  Number = 1:10,
  Die_1 = c(3, 2, 3, 1, 5, 3, 2, 2, 6, 4),
  Die_2 = c(3, 2, 5, 4, 2, 6, 1, 5, 3, 6)
)
# Exporting the dice table as dice.csv and dice.rds in the data folder
tw_csv_rds(dice_table, path = "data/dice")
```

tw_table()

```{r tw_table}
# creating a table with 5 rows per page
tw_table(pokemon, pagesize = 5)
```

tw_filter_select()

```{r tw_filter_select}
# Filtering on Type.1 = Fire and selecting the columns Name and Type.1
tw_filter_select(
  pokemon, 
  filter_col = "Type.1", 
  filter_value = "Fire", 
  select = c("Name", "Type.1")
  ) %>% head(5)
```
