# Analyzing Covid-19 data using parameters

### Selected parameters

covidData parameter: `r paste(params$covidData)`  
covidYear parameter: `r paste(params$covidYear)`  
covidMonth parameter: `r paste(params$covidMonth)`  

### Inspecting data

Pseudocode pyramid plot

- load the data
- select the right columns using the selected parameters
- change the geoId type to factor
- calculate the total cases and deaths per geoId  
- divide the cases and deaths per geoId by the geoId's population 
- take the log1p of total cases and total deaths so that they can be plotted in the same ggplot
- make pyramid plot out of the total cases and total deaths @alexAnswerSimplerPopulation2022; @eeeeedAnswerSimplerPopulation2019; @elfertsAnswerSimplerPopulation2013; @guyabelAnswerSimplerPopulation2016; @Top50Ggplot2

```{r}
# reading covidData parameter
covidData <- read.csv(params$covidData)

# filtering on selected parameters
covidDataFil <- covidData %>% dplyr::filter(year==params$covidYear & month==params$covidMonth) 

# inspect the selected tables
covidDataFil %>% reactable::reactable(defaultPageSize = 5, compact = TRUE, filterable = TRUE)
```

```{r}
# make geoId a factor
covidDataFil$geoId <- factor(covidDataFil$geoId, levels = unique(covidDataFil$geoId))

# check type
str(covidDataFil$geoId)

# select the columns geoId and population
cdGeoPop <- covidDataFil %>% select(geoId, countriesAndTerritories, popData2020) %>% unique()

# calulate total cases and deaths per geoId
totalCases <- aggregate(covidDataFil$cases, list(geoId = covidDataFil$geoId), sum)
names(totalCases)[2] <- "totalCases"
totalDeaths <- aggregate(covidDataFil$deaths, list(geoId = covidDataFil$geoId), sum)
names(totalDeaths)[2] <- "totalDeaths"

# merge dataframes together like left join would in SQL
covidDataJoined <- cdGeoPop %>% 
  left_join(totalCases, by="geoId") %>% 
  left_join(totalDeaths, by="geoId")

# normalize the data by dividing by population
covidDataJoined <- covidDataJoined %>% mutate(
    casesNormalized=totalCases / popData2020,
    deathsNormalized=totalDeaths / popData2020)

# death and cases in the same row in preparation for the plot (tidying data)
covidDataJoined <- covidDataJoined %>% pivot_longer(cols=c(totalCases,totalDeaths), names_to = "totalCases_totalDeaths", values_to = "Normalized")

# check the dataframe
covidDataJoined

# make totalCases_totalDeaths a factor
covidDataJoined$totalCases_totalDeaths <- factor(covidDataJoined$totalCases_totalDeaths, levels = unique(covidDataJoined$totalCases_totalDeaths))

# check type
str(covidDataJoined$totalCases_totalDeaths)

# make pyramid plot
# https://stackoverflow.com/questions/14680075/simpler-population-pyramid-in-ggplot2
library(ggplot2)

covidDataJoined %>%
  mutate(countriesAndTerritories_reordered = reorder(countriesAndTerritories, log1p(Normalized))) %>%
  ggplot(aes(
    x = log1p(Normalized) * ifelse(totalCases_totalDeaths == "totalDeaths", -1, 1), # if it's totalDeaths then multiply by -1 otherwise multiply by 1
    y = countriesAndTerritories_reordered,
    fill = totalCases_totalDeaths)) +
  geom_col() +
  labs(
    title = paste(
      "Impact of the corona pandemic as of", 
      month(params$covidMonth, label = TRUE, abbr = FALSE), 
      params$covidYear),
    x = "log1p((Cases | Deaths) / population)",
    y = "Countries & Territories") +
  theme(axis.text.y = element_text(size = 8)) # text size y axis

```



