# Analyzing Covid-19 data using parameters {#08}

Data source: @DataDailyNumber2022

### Selected parameters

covidData parameter: `r paste(params$covidData)`  
covidYear parameter: `r paste(params$covidYear)`  
covidMonth parameter: `r paste(params$covidMonth)`  

### Psuedocode pyramid plot

Pseudocode pyramid plot

- load the data
- select the right columns using the selected parameters
- change the geoId type to factor
- calculate the total cases and deaths per geoId  
- divide the cases and deaths per geoId by the geoId's population 
- take the log1p of total cases and total deaths so that they can be plotted in the same ggplot
- make pyramid plot out of the total cases and total deaths @alexAnswerSimplerPopulation2022; @eeeeedAnswerSimplerPopulation2019; @elfertsAnswerSimplerPopulation2013; @guyabelAnswerSimplerPopulation2016; @Top50Ggplot2

### Data reading

```{r read the covid cases data}
# reading covidData parameter
covidData <- read.csv(params$covidData)

# filtering on selected parameters
covidDataFil <- covidData %>% dplyr::filter(year==params$covidYear & month==params$covidMonth) 

# inspect the selected tables
covidDataFil %>% reactable::reactable(defaultPageSize = 5, compact = TRUE, filterable = TRUE)
```

### prepare the data for the pyramid plot

```{r prepare the data for the pyramid plot}
# make geoId a factor
covidDataFil$geoId <- factor(covidDataFil$geoId, levels = unique(covidDataFil$geoId))

# check type
str(covidDataFil$geoId)

# select the columns geoId and population
cdGeoPop <- covidDataFil %>% dplyr::select(geoId, countriesAndTerritories, popData2020) %>% unique()

# calulate total cases and deaths per geoId
totalCases <- aggregate(covidDataFil$cases, list(geoId = covidDataFil$geoId), sum)
names(totalCases)[2] <- "totalCases"
totalDeaths <- aggregate(covidDataFil$deaths, list(geoId = covidDataFil$geoId), sum)
names(totalDeaths)[2] <- "totalDeaths"

# merge dataframes together like left join would in SQL
covidDataJoined <- cdGeoPop %>% 
  left_join(totalCases, by="geoId") %>% 
  left_join(totalDeaths, by="geoId")

# normalize the data by dividing by population
covidDataJoined <- covidDataJoined %>% mutate(
    casesNormalized=totalCases / popData2020,
    deathsNormalized=totalDeaths / popData2020)

# death and cases in the same row in preparation for the plot (tidying data)
covidDataJoined <- covidDataJoined %>% pivot_longer(cols=c(totalCases,totalDeaths), names_to = "totalCases_totalDeaths", values_to = "Normalized")

# check the dataframe
covidDataJoined

# make totalCases_totalDeaths a factor
covidDataJoined$totalCases_totalDeaths <- factor(covidDataJoined$totalCases_totalDeaths, levels = unique(covidDataJoined$totalCases_totalDeaths))

# check type
str(covidDataJoined$totalCases_totalDeaths)
```

### Pyramid plot

```{r pyramid plot}
# make pyramid plot
# https://stackoverflow.com/questions/14680075/simpler-population-pyramid-in-ggplot2
library(ggplot2)

covidDataJoined %>%
  mutate(countriesAndTerritories_reordered = reorder(countriesAndTerritories, Normalized)) %>%
  ggplot(aes(
    # if it's totalDeaths then multiply by -1 otherwise multiply by 1. Log1p is used to avoid log(0)
    x = log1p(Normalized) * ifelse(totalCases_totalDeaths == "totalDeaths", -1, 1), 
    y = countriesAndTerritories_reordered,
    fill = totalCases_totalDeaths)) +
  geom_col() +
  labs(
    title = paste(
      "Impact of the corona pandemic as of", 
      month(params$covidMonth, label = TRUE, abbr = FALSE), 
      params$covidYear),
    x = "log1p((Cases | Deaths) / population2020)",
    y = "Countries & Territories") +
  theme(axis.text.y = element_text(size = 8)) + # text size y axis
  scale_color_brewer(palette = "Dark2") # colorblind friendly color scale
```

Pseudocode heatmap plot @raschkaShortTutorialDecent0000

- devide the cases and deaths by the population data of 2020 to normalize the data
- sort the columns by day number @bobbittHowSortData2021
- use pivor wider to give every day a column in preparation for the heatmap
- store the names of the countries and territories @tomasDropTRUEDoesn2014
- convert all day columns to numeric
- convert type to matrix and assign it the stored names
- plot the heatmap

### Preparing the data for the heatmap

Cases @raschkaShortTutorialDecent0000:

```{r prepare cases data for the heatmap}
# matrix preparation for cases
covidDataMatrixCases <- covidDataFil %>% mutate(Normalized = cases/popData2020) %>%  dplyr::select(countriesAndTerritories, day, Normalized) 

# sort dataframe so that the days are ordered from 1 to 30 or 31
covidDataMatrixCases <- covidDataMatrixCases[order(covidDataMatrixCases$day), ]

# column per day
covidDataMatrixCases <- covidDataMatrixCases %>% pivot_wider(names_from = day, values_from = Normalized)

# store the names because the matrix only wants one type (numerical, string, etc..)
# drop=TRUE simplifies the values to store it only as a vector and not as a column of a dataframe

rnames <- covidDataMatrixCases[,1, drop=TRUE]

# convert all columns except the first to numeric
covidDataMatrixCases <- data.matrix(covidDataMatrixCases[,2:ncol(covidDataMatrixCases)])

# assign column names to matrix
rownames(covidDataMatrixCases) <- rnames
```

Deaths @raschkaShortTutorialDecent0000:

```{r prepare deaths data for the heatmap}
# matrix preparation for cases
covidDataMatrixDeaths <- covidDataFil %>% mutate(Normalized = deaths/popData2020) %>%  dplyr::select(countriesAndTerritories, day, Normalized) 

# https://www.statology.org/sort-dataframe-by-column-in-r/
covidDataMatrixDeaths <- covidDataMatrixDeaths[order(covidDataMatrixDeaths$day), ]

# column per day
covidDataMatrixDeaths <- covidDataMatrixDeaths %>% pivot_wider(names_from = day, values_from = Normalized)

# store the names because the matrix only wants one type (numerical, string, etc..)
# drop=TRUE simplifies the values to store it only as a vector and not as a column of a dataframe

rnames <- covidDataMatrixDeaths[,1, drop=TRUE]

# convert all columns except the first to numeric
covidDataMatrixDeaths <- data.matrix(covidDataMatrixDeaths[,2:ncol(covidDataMatrixDeaths)])

# assign column names to matrix
rownames(covidDataMatrixDeaths) <- rnames
```

### Heatmap @raschkaShortTutorialDecent0000

```{r plot the cases heatmap, fig.width=10}
# heatmap plot cases
heatmap(
  covidDataMatrixCases, # data
  Rowv=NA, Colv=NA, # no clustering
  main = paste("Cases/population2020 per day as of", 
      month(params$covidMonth, label = TRUE, abbr = FALSE), 
      params$covidYear), # title name
  xlab = "Day", # x-axis name
  ylab = "Country or Territory" # y-axis name
  )
```

```{r plot the deaths heatmap, fig.width=10}
# heatmap plot deaths
heatmap(
  covidDataMatrixDeaths, # data
  Rowv=NA, Colv=NA, # no clustering
  main = paste("Deaths/population2020 per day as of", 
      month(params$covidMonth, label = TRUE, abbr = FALSE), 
      params$covidYear), # title name
  xlab = "Day", # x-axis name
  ylab = "Country or Territory" # y-axis name
  )
```
